---
title: "Regeneration Paper"
author: "micah gearhart"
date: "5/30/2015"
output: 
  md_document:
    variant: markdown_github
---

# Libraries
```{r libraries,eval=T}
library("downloader")
library("Biostrings")
library("ShortRead")
library("BiocParallel")
library("preprocessCore")
library("GenomicAlignments")
library("readxl")
library("dplyr")
library("ggplot2")
library("magrittr")
library("tidyr")
library("pryr")
library("readxl")
library("flashClust")
library("quantro")
library("stringdist")

#setwd("/mnt/afp/micah/R/umn-gcd-bioinformatics-axolotl/axo4")
setwd("/home/bardwell/gearhart/amex-miR/")
```

# Convert Original Data from Solexa
```{r solexa,eval=F}
#Illumina small RNA-Seq 3' adapter:  TCGTATGCCGTCTTCTGCTTGT
adapter <- DNAString("TCGTATGCCGTCTTCTGCTTGT")

am1<-readBaseQuality("/home/echeverr/gearhart/miRNA-Solexa/AM0001", seqPattern="*_1.seq.txt", prbPattern="*_1.prb.txt",type="Solexa")
length(am1)
am1<-am1[alphabetScore(am1) > 800]
am1<-am1[vcountPattern("-",sread(am1))==0]
am1trimmed<-trimLRPatterns(Rpattern=adapter,Lpattern=adapter,subject=am1,max.Rmismatch=0.1)
length(am1)
writeFastq(am1trimmed,file="am1_qf_trim.fastq",full=TRUE,compress=FALSE)

am2<-readBaseQuality("/home/echeverr/gearhart/miRNA-Solexa/AM0002", seqPattern="*_1.seq.txt", prbPattern="*_1.prb.txt",type="Solexa")
length(am2)
am2<-am2[alphabetScore(am2) > 800]
am2<-am2[vcountPattern("-",sread(am2)) == 0]
am2trimmed<-trimLRPatterns(Rpattern=adapter, Lpattern=adapter, subject=am2,max.Rmismatch=0.1)
length(am2)
writeFastq(am2trimmed,file="am2_qf_trim.fastq",full=TRUE,compress=FALSE)
```

# Build a bowtie index from mirBase 21 
```{r bowtie_build,eval=F}
#download("ftp://mirbase.org/pub/mirbase/21/mature.fa.gz",dest="mature.fa.gz")
#gunzip("mature.fa.gz")
mature<-readRNAStringSet("mature.fa")
matureDNA<-DNAStringSet(mature)
writeXStringSet(matureDNA,"matureDNA.fa")
system2(command="/home/bardwell/shared/bowtie-1.1.2/bowtie-build",
        args="-f matureDNA.fa mature",
        stdout="build_stderr.txt",stderr="build_stdout.txt")
```

# Map trimmed reads to mature index
```{r bowtie_map,eval=F}
#am1
system2(command="/home/bardwell/shared/bowtie-1.1.2/bowtie",
        args=paste0("-t -p 24 -e 99999 --norc mature ",
                    "--solexa-quals am1_qf_trim.fastq  ",
                    "--un am1_qf_trim.unmapped2mature.fastq ",
                    "-S am1_qf_trim.mature.sam"),
        stdout="am1_out.txt",stderr="am1_err.txt")
asBam("am1_qf_trim.mature.sam",destination="am1_qf_trim.unsorted")
sortBam("am1_qf_trim.unsorted.bam",destination="am1_qf_trim.mature")

#am2
system2(command="/home/bardwell/shared/bowtie-1.1.2/bowtie",
        args=paste0("-t -p 24 -e 99999 --norc mature ",
                    "--solexa-quals am2_qf_trim.fastq  ",
                    "--un am2_qf_trim.unmapped2mature.fastq ",
                    "-S am2_qf_trim.mature.sam"),
        stdout="am2_out.txt",stderr="am2_err.txt")
asBam("am2_qf_trim.mature.sam",destination="am2_qf_trim.unsorted")
sortBam("am2_qf_trim.unsorted.bam",destination="am2_qf_trim.mature")
```

# Read in sequences unmapped to MirBase 21
This section reads in the fastq sequences that did not have any matches to mirBase 21.  The fullSeqHash function builds a hash table and enumerates the number of times each sequence occured in each sample.

```{r readUnmapped,eval=F}
fls<-list.files(".",pattern=glob2rx("am?_qf_trim.unmapped2mature.fastq"),full.names=F)
fls
temp<-do.call(cbind,lapply(fls,readFastq))
table(width(sread(temp[[1]])))

fullSeqHash <- function (x) {
  x<-x[width(x)>19 & width(x)<24] #based on table above
  seed<-as.data.frame(table(as.factor(as.character(sread(x)))))
  seed<-seed[with(seed,order(-Freq)),]
  return(seed)
}

amFSH<-lapply(temp,fullSeqHash)
names(amFSH)<-c("am1","am2")
for (i in names(amFSH)) {amFSH[[i]]$sample <- i}
amFSH<-do.call(rbind,amFSH) %>% spread(sample,Freq,fill=0)
dim(amFSH)
amFSH[grep("GGCTTCAAGGATCGCTCGGG",amFSH$Var1),]
save(amFSH,file="amFSH.rdata")
```

# Cluster similar Sequences
```{r cluster,eval=F}
(nt<-detectCores())
register(MulticoreParam(workers=nt))
#load("amFSH.rdata")
amFSHs<-amFSH %>% filter((am1+am2)>7)
dim(amFSHs)
system.time(d<-as.dist(stringdistmatrix(amFSHs$Var1,amFSHs$Var1,method="lv",nthread=nt)))
object_size(d)
system.time(h<-hclust(d))
object_size(h)
amFSHs$clust2<-cutree(h,h=2)
amFSHs$clust3<-cutree(h,h=3)
amFSHs$clust4<-cutree(h,h=4)
amFSHs$clust5<-cutree(h,h=5)
amFSHs$clust6<-cutree(h,h=6)
amFSHs$clust7<-cutree(h,h=7)
amFSHs$clust8<-cutree(h,h=8)
save(amFSHs,file="amFSHs.rdata")
```

# Analysis of miRbase 21 Mapping Results
This section counts the number of reads in the bam files that are mapped to each miR.  A summary of the known changes or changes that make up more than 5% of the total are tabulated.
```{r countBam,eval=F}
(fls <- list.files(".", pattern=glob2rx("am*qf_trim.mature.bam"),full=TRUE))
bamlst <- BamFileList(fls,yieldSize=1000000)
(gr <- GRanges(seqnames(seqinfo(bamlst[[1]])),IRanges(1, seqlengths(bamlst[[1]]))))

#countBam(bamlst[[1]],param=ScanBamParam(flag=scanBamFlag(isUnmappedQuery = TRUE,isNotPassingQualityControls=NA)))

am1_mat<-countBam(bamlst[[1]], param=ScanBamParam(which = gr))
am2_mat<-countBam(bamlst[[2]], param=ScanBamParam(which = gr))
am1_mat<-select(am1_mat,space,records)
colnames(am1_mat)<-c("miR","am1")
am2_mat<-select(am2_mat,space,records)
colnames(am2_mat)<-c("miR","am2")
am_mature<-merge(am1_mat,am2_mat,by="miR")
row.names(am_mature)<-am_mature[,"miR"]
am_mature<-am_mature[,-1]

#check distribution
matdensity(as.matrix(log2(am_mature[am_mature$am1+am_mature$am2 > 7,]+0.5)))

#add family
am_mature$family<-sapply(strsplit(row.names(am_mature),"-"), function(x) paste0(x[2],"-",gsub("\\*","",x[3])))
head(am_mature)

#Add unmapped reads to am_mature
load("amFSH.rdata")
unknowns<-amFSH
row.names(unknowns)<-unknowns[,"Var1"]
unknowns<-unknowns[,-1]
unknowns$family<-"unknown"
head(unknowns)

#am_mature<-rbind(am_mature,
#                 data.frame(am1=as.integer(sum(amFSH[,"am1"])),
#                            am2=as.integer(sum(amFSH[,"am2"])),
#                            row.names="unknown",
#                            stringsAsFactors = F))


am_mature<-rbind(am_mature,unknowns)
#check density with unknowns
matdensity(as.matrix(log2(am_mature[am_mature$am1+am_mature$am2 > 5,c("am1","am2")]+0.5)))
#qtest <- quantro(as.matrix(log2(am_mature[am_mature$am1+am_mature$am2 > 7,c("am1","am2")]+0.5)),
#groupFactor = factor(c("am1","am2")))

#subset to samples with more than 7 counts
am_mature<-am_mature[am_mature$am1+am_mature$am2 > 5,c("family","am1","am2")]

#qqnorm(log2(am_mature$am1+0.5))
#qqnorm(log2(am_mature$am2+0.5))
#not norml

#quantile normalize
#am_mature<-cbind(am_mature,
#                 as.data.frame(normalize.quantiles(as.matrix(am_mature[,c("am1","am2")]))))
#
#                 varianceStabilizingTransformation(as.matrix(am_mature[,c("am1","am2")])))

#colnames(am_mature)<-c("family","am1","am2","qn.am1","qn.am2","vst.am1","vst.am2")
#colnames(am_mature)<-c("am1","am2","family","qn.am1","qn.am2")

#am_mature$log2am1<-log2(am_mature$qn.am1+0.5)
#am_mature$log2am2<-log2(am_mature$qn.am2+0.5)
#qqnorm(am_mature$rl.am1)
#m<-am_mature[am_mature$family!="unknown",]$log2am1-am_mature[am_mature$family!="unknown",]$log2am2
#a<-(am_mature[am_mature$family!="unknown",]$log2am1+am_mature[am_mature$family!="unknown",]$log2am2)/2
#m<-am_mature$vst.am1-am_mature$vst.am2
#a<-(am_mature$vst.am1+am_mature$vst.am2)/2
#plot(a,m,cex=0.4,pch=16,col=ifelse(am_mature$family=="unknown","red","black"))
#abline(h=0,col="black")
#am_mature$logfc<-am_mature$vst.am2-am_mature$vst.am1
#head(am_mature)

#tbl1_miRs<-c("let-7","let-7a","let-7f","miR-10","miR-21",
#             "miR-27","miR-128","miR-140","miR-196","miR-199a","miR-206")
#tbl1<-am_mature %>% dplyr::filter(family %in% tbl1_miRs) %>% group_by(family) %>% 
#  summarize(control=sum(am1),blastema=sum(am2)) %>% 
#  mutate(fam=factor(family,levels=tbl1_miRs)) %>% 
#  select(fam,control,blastema)
#tbl1
```

# Export Tables for Known Sequences

```{r export_known_tables,eval=F}
temp<-am_mature
colnames(temp)<-c("miR Family","Control Counts","Blastema Counts",
                  "Normalized Control Counts","Normalized Blastema Counts",
                  "Log2 Control Counts","Log2 Blastema Counts","Log2 Fold Change Blastema/Control")
write.csv(temp,"Supplemental_Table_1.csv",quote=F)

temp2<-tbl1
colnames(temp2)<-c("miR Family","Control Counts","Blastema Counts")
write.csv(temp2,"Table_1.csv",quote=F)
```

# Figure 1a - Barchart of mirBase 21 Mapping Data 
Create a graphical summary of known sequences that map to mirBase 21.  This graphic shows sequences that are present at > 5% abundance in either sample.

```{r barchart,eval=F,fig.height=8,fig.width=8}
#Convert counts to fractions for each family
asFamilies<-am_mature %>% group_by(family) %>% 
  summarize(control=sum(am1),blastema=sum(am2)) %>%
  mutate(control = control / sum(control),blastema=blastema/sum(blastema)) 

#Rename "unknown" to "Uncategorized" outside of dplyr for grouping
asFamilies[asFamilies$family=="unknown","family"]<-"Uncategorized"

asFamilies <- asFamilies %>% 
  #combined Abundance
  mutate(family = ifelse((control + blastema) < 0.05,"Combined miRs with < 5% Abundance",family)) %>%
  mutate(family = factor(family, levels = rev(c("miR-1","miR-1a","miR-1b",
                                         "let-7","let-7a","let-7f",
                                         "miR-21","miR-140","miR-199a","miR-206",
                                         "Combined miRs with < 5% Abundance",
                                         "Uncategorized"))))   
  
  #Alternatively use Sample Specific Abundance
  #mutate(family = ifelse(!(control > 0.05 | blastema > 0.05),"Combined miRs with < 5% Abundance",family)) 
  
#set up colors based on factor levels
colors<-c("#cbc9e2","#9e9ac8","#6a51a3",
          "#fdbe85","#fd8d3c","#d94701",
          "#56B4E9", "#009E73","#F0E442", "#0072B2", 
          "gray","black")
          
asFamilies %>% 
  gather(sample,reads,control,blastema) %>% dplyr::filter(reads!=0) %>%  
  ggplot(aes(x=sample,y=reads,fill=family,order=family)) +
    scale_fill_manual(values=rev(colors),guide = guide_legend(reverse=TRUE)) +
    geom_bar(stat="identity") +theme_classic()
```

# Create Table of Validated Novel miRs
```{r validated_miRs,eval=F}
load("amFSHs.rdata")
head(amFSHs)
novel_mirs<-amFSHs %>% 
  mutate(Var1=as.character(Var1)) %>% 
  group_by(clust4) %>% 
  summarize(am1s=sum(am1),am2s=sum(am2),
            seq=list(as.character(Var1)),
            MaxSeq=as.character(Var1)[which.max(am2)]) 
  
validated_mirs <-c("CCACGCAGGGATGTGGCTTT","GGCTTCAAGGATCGCTCGGGCA","GGCCGCAAAGGTTGCACAGGG",
                   "ACAAACCATCTTGGGCTGTTAC","CCCCGACGGGGAGGGCAATGTTT", "TCCTCCGAGCCAAGCTTTTG",
                   "ACATATCTACAGTCCTACTATA","CCGAGTTCTGCATTCCGCTTGAA","CCCACGCAGGGATGTGGCTTA",
                   "AGCCGCATTGTACCGGGCTAT","CGCCGCATTGTACCGGGCTCT","CTTTCCGTCGGATGCTTCCAGC")
#"Mon Jun 15 10:05:47 2015"
validated_mirs<-c("CCACGCAGGGATGTGGCTTTT","GGCTTCAAGGATCGCTCGGGC","CCCCGACGGGGAGACTCACTTTT",
                  "TCCTCCGAGCCAAGCTTTTG","AGCCGCATTGTACAGGGCTAT","AGCCGCATTGTACAGGGCTAT",
                  "CTTTCCGTCGGATGCTTACAGC")

#Tue Jun 16 11:41:02 2015 - Original Sequences from Karen's Novel miRs Table
validated_mirs<-c("CCACGCAGGGATGTGGCTTT","GGCTTCAAGGATCGCTCGGGCA","CCCCGACGGGGAGGGCAATGTTT",
                  "AGCCGCATTGTACCGGGCTAT","CGCCGCATTGTACCGGGCTCT","CTTTCCGTCGGATGCTTCCAGC")
names(validated_mirs)<-c("KEM42017","KEM68162","KEM43147","KEM21802","KEM48467","KEM56768")

seqX<-lapply(novel_mirs$seq,unlist)

# tbl1$cluster<-as.numeric(sapply(seq_along(validated_mirs),function(x) grep(validated_mirs[x],seqX)))

#Results Table
validated_mirs<-validated_mirs[!names(validated_mirs) %in% "KEM21802"]
tbl1<-as.data.frame(validated_mirs)

#pre-allocate
tbl1$num_seqs<-0
tbl1$cluster<-0
tbl1$seqs<-"none"

#for loop to fill table
for (i in 1:nrow(tbl1)) {
  tbl1[i,"cluster"]<-which(unlist(sapply(seq_along(seqX),function(x) validated_mirs[i] %in% seqX[[x]])))
  tbl1[i,"num_seqs"]<-length(unlist(novel_mirs[tbl1[i,"cluster"],"seq"]))
#  tbl1[i,"seqs"]<-paste0(as.character(unlist(novel_mirs[tbl1[i,"cluster"],"seq"])),collapse=",")
}
tbl1$am1<-novel_mirs[tbl1$cluster,"am1s"]
tbl1$am2<-novel_mirs[tbl1$cluster,"am2s"]
tbl1$MaxSeq<-novel_mirs[tbl1$cluster,"MaxSeq"]
 tbl1<-as.data.frame(tbl1)
 tbl1<-tbl1[tbl1$num_seqs > 0,]


  #Use this line to report all sequences in cluster
  #summarize(am1s=sum(am1),am2s=sum(am2),seq=list(as.character(Var1)))

#Make a Table for the subset of MiRs characterized in this paper

# this one got filtered out somehow?
#validated_mirs<-"CTTTCCGTCGGATGCTTACAGC"

#f<-function(i) { lapply(blah[i]$seq,function(x) grep(novel_mirs[1],x)) }
#lapply(seq_along(blah$seq), f)
seqX<-lapply(novel_mirs$seq,unlist)
#novel_mirs[unlist(sapply(seq_along(validated_mirs),function(x) grep(validated_mirs[x],seqX))),]

#amFSHs[grep("CTTTCCGTCGGATGCTTACAGC",amFSHs$Seed),]
#stringDist("CTTTCCGTCGGATGCTTACAGC","TCGTATGCCGTCTT")

#sapply(seq_along(validated_mirs),function(x) grep(validated_mirs[x],seqX))


```

# Write out all Novel Families and blast max.sequences against mirBase to find closest known miRs
```{r export_novel,eval=F}
nm<-DNAStringSet(novel_mirs$MaxSeq)
names(nm)<-paste0("axo-",novel_mirs$clust4)
writeXStringSet(nm,"nm.fa")

#vm<-DNAStringSet(validated_mirs)
#writeXStringSet(vm,"vm.fa")

#cmd<-"blastn -query vm.fa -outfmt 10 -db ../miRs/mirBase/matureBLAST -word_size 5  "
system("blastn -query nm.fa -outfmt 10 -out nm_matureBlast.csv -db ../miRs/mirBase/matureBLAST -word_size 4")

br<- read.csv("nm_matureBlast.csv",stringsAsFactors=F,header=F)
colnames(br)<-c("qseqid","sseqid","pident","length","mismatch","gapopen","qstart","qend","sstart","send" ,"evalue","bitscore")

#set threshold on blast hits
br<-br[br$evalue < 1 & (br$length-br$mismatch-br$gapopen) > 15,]
dim(br)
length(unique(br$qseqid))/length(nm)

```


#### SESSION INFO
```{r sessionInfo,eval=T}
sessionInfo()
```

